The Decline in U.S Birthrates in Recent Years is Indicative of Socio-economic Changes

### 2022-02-28 Output/paper 2.Rmd
Last changes to code and writing

---
title: "Decline in U.S Birthrates is Indicative of Socio-economic Changes"
author: "Olaedo Okpareke, Arsh Lakhanpal, Swarnadeep Chattopadhyay"
thanks: 'Code, data and replication are available at: https://github.com/OlaedoOkpareke/Paper2Repository and https://doi.org/10.48152/ssrp-8bs7-7e32' 
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Birth rates provide useful information on population growth, and an above-replacement birth rate is indicative of stability. However, previous research has found that birth rates in industrialized societies such as the U.S have been decreasing in recent years. We use data from the American Economic Association to analyze this by using graphs and tables to observe the trend in birth rates of different demographics, using R [@citeR] and other packages. We also observe the trends behind changing economic and cultural factors that could affect the birth rates. We find that birth rates of young women of different races have been on a steep decline over the past 20 years, and factors that are known to decrease birth rates have increased. This report shows that the decline in birth rates are a symptom of cultural and economic changes between two generations of women. While this indicates that the U.S has gotten more culturally liberated, the declining birth rates shows the systemic effects of the worsening economy and increasing costs on the average american woman."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references2.bib

@@

A nation's birthrate for a particular year is the total number of live births per 1000 people for that year. In order to measure the birth rates and fertility levels to forecast population growth for the United States, the National Center for Health Statistics (NCHS) aggregates data on birth rates for women by age, race, educational status, and other factors. The total population of the country is the total number of citizens and other residents in the country, regardless of legal status. It is used in many types of planning, such as planning and decisions on public services, health care, education, government funding, transportation, and politics. The total population is obtained through a population census, provided by the CDC Surveillance, Epidemiolody, and End results program (CDC SEER). This information also breaks down the population by age, race, and socio-economic factors. In this report,we are interested in the trend in birth rates of young women from 2000 to 2020 as well as the breakdown of these birth rates by the race and age demographics of the mother. We are also interested in the relationship between the change in economic, social and cultural aspects of the United states and the change in these birth rates.

We obtained the dataset from the American Economic Association [@citewebsite]. We loaded in, cleaned and analyzed the data using  `R` [@citeR], `dplyr`, [@citedplyr], `tidyr` [@citetidyr],`haven` [@citehaven], and `tidyverse` [@citetidyverse] packages. Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr], `usmap` [@citeusmap], `gridExtra` [@citegridExtra] and `kableExtra` [@citekable]. We first created histograms and scatter plots which showed the most important variables. We then created line plots and a map which showed the trend in birth rates over the years, for different age groups, races, and states. We then created tables which showed the relative contributions of our target demographics on the change in birth rates and the differences in their populations, as well as showed the change in values for possible economic and societal factors that may be correlated to birth rates.

Young women are in no rush to have children, as this report shows that birth rates have declined for the 15-24 age groups over the past 20 years. The report also shows that birth rates have declined for all races analyzed over the same period of time. This implies that this decline is a general national issue attributed to national factors, as opposed to factors that mainly affect a specific demographic. The birth rates for our target demographics have declined to the point that they make up a rather large contribution to the total decline in US birth rates. Birth rates for our age demographics have experienced large decreases in some states, while those in other states have had a relatively smaller decrease or no decrease at all. This may be due to the fact that some economic or social changes of the country affect some geographic areas more than others. These economic and social factors suspected to have an influence on birth rates also changed over the past 20 years in ways that would correlate with declining birth rates. According to [@citeadsera], these rates of birth decline are similar to declines in other high income, developed countries.

While this census is widely used to forecast population growth and make important decisions on the future of the country, biases and problems in data collection could paint an incorrect view of birth rates and the total population. The declining birth rates also brings to light the problems of increasing cost of living, and the increasingly expensive opportunity costs of working and childcare for women throughout the country [@citebarry]. This is problematic as although positive cultural changes have led to more freedom for women than before, as well as decreased teenage births, this also comes with the fact that many are no longer able to afford children without a strong financial background.The data does not collect information in certain groups of people, thus removing them and their experiences from the ranks, further propagating their discrimination in society. The data collection also helps propagate preconceived biases against young women living in poor, low education areas as well as women of color as being irresponsible with their bodies. This is because the data only collects the statistics but does not provide meaning, causes, and explanations behind these statistics [@Gold2001].  There are also concerns about biases in data collection of data, political impacts of the data as well as the future state of population growth in the US and its consequences.

@@

The datasets for this report were gotten from the data section for "The Puzzle of Falling US Birth Rates since the Great Recession" [@citepaper], a paper from the Journal of Economic Perspectives [@citewebsite]. The data used by the paper was compiled by various sources such as the NCHS Natality Database [@citenchs], CDC SEER [@citecdcseer],the Current Population Survey Annual Social and Economic Supplement (CPS ASEC) , the New York Federal Reserve [@citefed], and the Integrated Public Use Microdata Series (IPUMS) [@citeipums]. The paper and datasets were last accessed on February 16th, 2022. The first two sources collect and estimate data via national census from US citizens and residents by county or state over the years 1968 to 2019. The birth rate data was gotten from a sample of 100 percent of birth certificates (i.e the population) in some states, and a sample of 50 percent of the birth certificate in other states, and this was randomly sampled [@citenchs]. The national population data comes from the census, a long running survey funded by the government released on a predetermined cycle, and counts the entire population of the United States by county, thus response was mandatory by government mandate [@citecdcseer]. The datasets from the other sources compiled by the paper authors typically include data by year and by state for several demographics. These datasets were gotten from surveys so not everyone responded, thus the data was often recorded as 0, `NA`, or estimated.The data is mainly numeric, counting the number of births, total population, or women in a particular economic/social/educational demographic, age group, or ethnicity. Meanwhile, categorical data mainly appeared in the form of state.

@@

For data cleaning, our target sample and focus was young mothers from two generations: millennials, who were between 15 to 24 years in 2000, and generation Z, who were between 15 to 24 years old in 2020. We observed birthrates from 2000 to 2020, in order to see the effect of relatively short-term effects on birthrates between the generations. Millennials are often considered to be born between 1980 to 1995, and generation Z from 1996 to 2012 [@citedimock]. We also considered birth rates for White, Black, and Hispanic women that were in our target age ranges as we wanted to compare the birth rates between them. We thus filtered our datasets based on those criteria or used the closest criteria when they were not available, i.e. 2019 data is used if 2020 data is not available. We dropped any `NA` values from the dataset if they appeared.

@@

We included datasets and variables if they were considered important for answering the question behind birthrates or revealed important information. We included variables on race, as this was needed to compare birth rates for different races. We also had variables on total number of births and birth rates, both of which were used to analyze the decline in births. We had a total population variable, used to get the birth rates and our demographics' share of the population. We had variables on birth rates for different age groups, states, and years, all of which were used to analyze the trends in birth rates for different groups.  We also included variables on government policy, rent costs, student debt, religious beliefs, and childcare costs. These variables were used to signify a possible relationship between them and birth rates. All the variables listed were numeric, with the exception of race and age group.

We created several of our own datasets by modifying and joining variables. One dataset was gotten by adding all the educational statuses for each age-race combination from another dataset to get the total birth rate for each combination, and then compiling the values by state and year. We also got a variable for total population by adding the total population for each state in a year.  We obtained the final dataset used to create the map of birth rate decline by combining various other datasets by year, and subtracting the birth rates of the later years by that of the earlier years. The final dataset that showed how our demographics of interest contributed to total birth rate decline was created by joining several datasets all which compared important information, then dividing the columns of our demographic by the column of total birth rates to get the relative contributions. The final dataset used to display the change in explanatory variables over time was gotten by combining the datasets, each of which explained the changes in a single variable. Most final datasets include either 51 observations (for each state), 102 observations (for each state by year), or 20 observations (for each year).

We modified many variables for the report. We created variables for total birth rates of each race-age combination by adding up all educational statuses for each race and age group combination and dividing by the total population, as each person whose birth rates was recorded also had their educational status recorded. We also modified age group variables by changing the columns of 15-19 and 20-24 age to rows within one column. We did the same for race variables, modifying the columns to be rows within one group, and then multiplying the values column by 10,000 and dividing by the total female population to get the birth rates for each race. These variables were used to get the trend in birthrates for our age demographics, and trend in birthrates in race for our age demographics. We obtained a variable which got the decline in birth rates from 2000 to 2019 by state, by creating dividing the number of births by population to get birth rates, grouping them by year and state, and then subtracting the values of the later year from the values of the earlier year.

We created a variable for the total number of births by adding up all the births by state and then grouping them by year. We also created a variable that stated the total population of each race-age combination by similar methods. We obtained a variable that got the percentage of population of a demographic by dividing the sum of that demographic by the total population and multiplying by 100. We produced a variable on contribution to total birth rate decline by subtracting 2019 birth rates of the age group from the 2000 birthrates of the same age group and dividing that by the total birthrate value for 2000, then multiplying by 200. We got the variables for all our explanatory factors by grouping them by year and getting the mean values from all the states. For the average debt, we multiplied by 10,000 as that is how it was shown in the original source. We got the variable on the importance of religion by dividing the number of people who felt religion was important over the total number surveys. Note that for these variables we needed to use 2004 as 2000 was not available.

@@

We see that for the 15 to 19 age group, the data is quasi-normal, with most births for the relevant years being between 60 to 190 thousand. The data seems evenly skewed and unimodal. We see that teenage births tend to be less popular, as the data seems slightly right skewed. This also means that the average number of births will be below the median number of births, which makes sense as teen births in some areas of the US are far higher than others [@Kearney2012]. We also see that black teenage mothers are less common, as they have the lowest births for most years. This is most likely because they are a much smaller segment of the population than the others. However we see that black mothers still do manage to have a moderate amount of births, indicating they are disproportionately appearing as teenage mothers. Hispanic teenage mothers are more middle ground, having between 60 to 100 thousand births per year. White teenage mothers are the most common, due to them being the largest in population. They tend to have most of their distribution towards the higher numbers.

For the 20-24 age group, there seems to be a slight bimodal distribution, with many births being on two extremes. Due to this it is difficult to determine whether the mean or the median of births will be larger.  We see that there are far more women giving birth at this age group than in the previous age group. We see that black mothers are overwhelmingly on the lower end of births, which brings into question why teenage black mothers had more prevalent births compared to their elder counterparts. Hispanic mothers also tend to be at the lower end of the scale, but are more spread out. White births are the largest, taking the higher end of the spectrum, indicating their higher populations, or that white women are more likely to have children at these ages.We see that births for black and hispanic mothers at age 20-24 are only double the births of their teenage counterparts, while white mothers tend to be as much as 5 times the number of their teenage counterparts, Therefore, teenage motherhood in black and hispanic women may be more common.

In contrast, in (Figure \@ref(fig:figthree)) the distribution for the total population of women for the past 20 years seems uniformal, with almost all years having their own value. This is because the population of a nation either increases or decreases, rarely staying the same. Due to this we can say that the mean and median population are approximately the same. However there are some years which had about the same population, this could be due to declining population as a later year has the same population as an earlier year. From the scatterplot we see that the total female population has only been increasing in recent years, with a seemingly exponential increase. This increase in the population could be due to increased amounts of immigration and the increased lifespan of the elderly to offset the decreasing birthrates. However, the two dips in the scatterplot correspond to the previous histogram of which had years with similar values.

@@

(Figure \@ref(fig:figfour)) shows the trend of births for the two age groups in a 20 year period. We see that there has been a significant decrease in the birth rates, especially in the 15-19 age group, dropping to half of its 2000 levels. This indicates a decrease in teenage pregnancy nationwide. The birth rates for the 20-24 age dropped by around 40% over the same time period. This may be because generation Z women are deciding to have children later in life. There was a baby boom in 2006 [@citecatholic], which is the reason for the slight increase in birth rates around that time. However overall birth rates for young women has dropped.  According to [@Caldwell2006] there is little reason to believe these birth rates will increase any time soon. There are some possible explanations behind this phenomenon, including better sex education for teenagers leading to a decrease in teen pregnancy, more women focusing on their careers and thus postponing births [@citeexecutive], or the increased acceptance of abortions. These declining birth rates could also be the cause of declining population. There are also economic factors correlated to this decline, including the rising cost of living.

```{r figfour, out.width="90%", fig.cap = "Trend in Birth rates for the 15-19 and 20-24 age groups from 2000 to 2020", echo = FALSE}

@@

(Figure \@ref(fig:figfive)) shows the trend of births for different races of women from the ages of 15 to 24 over the past 20 years. We see that the birthrate for Hispanic mothers significantly decreased, going from about 63 births per 1000 population to just about 40 births for 1000 population. This can either be because the population of hispanic women has increased due to immigration, or that the number of births has decreased. The birth rates for black women has also halved, going from 50 births per 1000 to just 25 births per 1000. The white birth rates however has taken the largest decrease, going from about 125 births to about 65 births through the 20 years. We see that while millennial mothers (mothers in 2000) had a wide spread in the birth rates for different races, for generation Z they have converged more closely due to the decrease in birth rates for all races. This could be due to economic factors directly related to race, such as inequality in the health care system [@Davis2018] . As the income for hispanic families increases and they become more westernized, it is possible that they may be adhering to the western standard of fewer children [@immigbirth].

```{r figfive, out.width="90%", fig.cap = "Trend in Birth rates by Race from 2000 to 2020", echo = FALSE}

@@

(Table \@ref(tab:tabone)) shows the contribution of our main groups to the total decline in birth rates from 2000 to 2019. It also shows each group's share of the total female population for those years. We see that white women in both age groups have contributed the most to the decline, with a total of 18.2% of the total decrease. White 15 to 19 year old women have caused a 7% drop in the total birth rate, indicating a rapid decline in the teenage pregnancy for this race [@Wingo2011]. White 20 to 24 year old women have contributed the most to the decline, possibly due to economic and cultural factors over the past 20 years [@Maher2007]. Black women for both age groups have not shown a significant contribution to the decline in birth rates over this period however, either because teenage pregnancy for this rates have not decreased significantly or because they are a much lower percentage of the population. Black women are disproportionately poorer [@citepovertyrace], and low economic status leads to higher teenage pregnancy rates [@Gold2001]. The contribution of Hispanic women of ages 20 to 24 is less than their teenage counterparts, showing that there is not a large decrease in their birth rates as opposed to white women. They are also more likely to suffer from disproportional poverty. We see that white womens' share of the population from 2000 to 2019 has decreased due to the large decrease in birth rates. However black and hispanic womens' share of the population over this time period has either increased or not changed, likely due to increasing immigration from these two races over the past 20 years.

@@

# getting each race - age combination's total contribution to declining birth rates
contribution = birthrates %>%
  pivot_wider(names_from = year, values_from = birthratevalue) 
contribution$contribution = ((contribution$"2000" - contribution$"2019")/65.9)*200

# getting the percentage of the population
percentpop = share_of_pop %>%
  pivot_wider(names_from = year, values_from = value)
percentpop$"2000" = percentpop$"2000"*100
percentpop$"2019"= percentpop$"2019"*100

@@

(Figure \@ref(fig:figsix)) shows the number of births for those aged 15-24 for every from 2000 to 2019, according to the year's unemployment rate. We notice that the graph generally has a decreasing pattern in terms of the average number of births per year in the United States which is in line with what we observed earlier. Between 2004 and 2005, the birth rate for the 15-19 group continued to decrease whereas the birth rate for the 20-24 age group began to increase. We see that the unemployment rate is low in this period. According to [@citecatholic], in 2006, the United States had their highest number of births than in the last 45 years prior, causing a baby boom. However after this, the birth rate began to quickly decline and unemployment rates were high, reaching 8% by 2008 and continuing until 2012. This is likely due to the great recession that occurred during this period, such that it is possible women could not afford bringing new children. The birth rate, while still decreasing for both age groups, becomes less severe from 2012 onward, and we see that unemployment reduces. The initial conclusion made by this is that low unemployment rates are related to a higher number of births and higher unemployment rates are related to a lower number of births. However beyond 2014, despite the low unemployment rate, we see the birth rate continues to decrease. We note that the unemployment rate is more likely to affect women in the 20-24 age group compared to their younger counterparts, as they are more involved in the workforce.

```{r figsix, out.width="90%", fig.cap = "Average Birth Rates Vs level of Unemployment for the 15 to 19 and 20 to 24 age groups from 2000 to 2020", echo = FALSE}

@@

# creating scatterplot for the 15 to 19 age group
g1519 <- graph %>% ggplot(mapping = aes(x = year, y = brate_1519, color = Unemployment)) + geom_point() + theme_minimal() + scale_x_continuous(breaks = seq(1980,2020,by = 2))+ labs(x = "Year", y = "Birth Rate for Ages 15-19") + 
  labs(color= "Unemployment Rate (%)")

# creating scatterplot for the 20 to 24 age group
g2024 <- graph %>% ggplot(mapping = aes(x = year, y = brate_2024, color = Unemployment)) + geom_point() + theme_minimal() + scale_x_continuous(breaks = seq(1980,2020,by = 2))+ labs(x= "Year", y = "Birth Rate for Ages 20-24") +
  labs(color= "Unemployment Rate (%)")

@@

(Figure \@ref(fig:figseven)) shows the percentage decline in birth rates for young women, specifically the the 20-34 age group, by state from 2000 to 2019. We see that throughout the U.S there has been a decline in birth rates in most states, with some states having a more severe decline than others. We see that states on the west coast tend to have a larger decline than states on the east coast, with California having a very high decline in birth rates, and North Dakota having relatively little change. This could be attributed to the west coast being more liberal [@citeleft] and thus less adhering  gender roles, such that women have children later or not at all. In comparison the east coast, midwestern and southern states have less severe changes in their birth rates, with some having no change or a slight increase in birth rates. Coincidentally, these states tend to be more conservative or religious, so they accept gender norms and refusal of new changes in culture. The exception seems to be Utah, which is conservative and religious due to Mormonism but has the largest decline in birth rates. However, the general trend still holds, and as some states are more costly to live in than others, this could also explain the differences.

@@

# mean rent by year, actually recorded monthly
renttable = rent %>%
  group_by(year) %>%
  summarize(meanrent = mean(avg_grossrent))

>>>>

---
title: "The Decline in U.S Birthrates in Recent Years is Indicative of Socio-economic Changes"
author: "Olaedo Okpareke, Arsh Lakhanpal, Swarnadeep Chattopadhyay"
thanks: 'Code, data and replication are available at: https://github.com/OlaedoOkpareke/Paper2Repository and https://doi.org/10.48152/ssrp-8bs7-7e32' 
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Birth rates provide useful information on population growth, and an above-replacement birth rate is indicative of stability. However, previous research has found that birth rates in industrialized societies such as the U.S have been decreasing in recent years. We use data from the American Economic Association to analyze this by using graphs and tables to observe the trend in birth rates of different demographics, using R [@citeR] and other packages. We also observe the trends behind changing economic and cultural factors that could affect the birth rates. We find that birth rates of young women of different races have been on a steep decline over the past 20 years, and factors that are known to decrease birth rates have increased. This report shows that the decline in birth rates is the result of cultural and economic changes between two generations of women. While this indicates that the U.S has become more culturally liberated, the declining birth rates showcases systemic effects of the worsening economy and increasing costs on the average American woman."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references2.bib

@@

A nation's birth rate for a particular year is the total number of live births per 1000 people for that year. In order to measure the birth rates and fertility levels to forecast population growth for the United States, the National Center for Health Statistics (NCHS) aggregates data on birth rates for women by age, race, educational status, and other factors. The total population of the country is the total number of citizens and other residents in the country, regardless of legal status. It is used in many types of planning, such as planning and decisions on public services, health care, education, government funding, transportation, and politics. The total population is obtained through a population census, provided by the CDC Surveillance, Epidemiology, and End results program (CDC SEER). This information also breaks down the population by age, race, and socio-economic factors. In this report,we are interested in the trend in birth rates of young women from 2000 to 2020 as well as the breakdown of these birth rates by the race and age demographics of the mother. We are also interested in the relationship between the change in economic, social and cultural aspects of the United States and the change in these birth rates.

We obtained the dataset from the American Economic Association [@citewebsite]. We loaded in, cleaned and analyzed the data using  `R` [@citeR], `dplyr` [@citedplyr], `tidyr` [@citetidyr], `haven` [@citehaven], and `tidyverse` [@citetidyverse] packages. Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr], `usmap` [@citeusmap], `gridExtra` [@citegridExtra] and `kableExtra` [@citekable]. We first created histograms and scatter plots which showed the most important variables. We then created line plots and a map which showed the trend in birth rates over the years, for different age groups, races, and states. We then created tables which showed the relative contributions of our target demographics on the change in birth rates and the differences in their populations, as well as showed the change in values for possible economic and societal factors that may be correlated to birth rates.

Young women are in no rush to have children, as this report shows that birth rates have declined for the 15-24 age groups over the past 20 years. The report also shows that birth rates have declined for all races analyzed over the same period. This implies that this decline is a general national issue attributed to national factors, as opposed to factors that mainly affect a specific demographic. The birth rates for our target demographics have declined to the point that they make up a rather large contribution to the total decline in US birth rates. Birth rates for our age demographics have experienced large decreases in some states, while those in other states have had a relatively smaller decrease or no decrease at all. This may be because some economic or social changes of the country affect some geographic areas more than others. These economic and social factors that are suspected to have an influence on birth rates have also changed over the past 20 years in ways that would correlate with declining birth rates. According to the study in [@citeadsera], this decline in birth rates are similar to declines in other high income, developed countries.

While this census is widely used to forecast population growth and make important decisions on the future of the country, biases and problems in data collection could paint an incorrect view of birth rates and the total population. The declining birth rates also brings light to the problems of increasing cost of living, and the increasingly expensive opportunity costs of working and childcare for women throughout the country [@citebarry]. This is problematic as although positive cultural changes have led to more freedom for women than before and decreased teenage births,now  many women are no longer able to afford children without a strong financial background. The data does not collect information in certain groups of people, thus removing them and their experiences from the ranks, further propagating their discrimination in society. The data collection also helps propagate preconceived biases against young women living in poor, low education areas as well as women of color as being irresponsible with their bodies. This is because the data only collects the statistics but does not provide meaning, causes, and explanations behind these statistics [@Gold2001].  There are also concerns about biases in data collection of data, political impacts of the data as well as the future state of population growth in the US and its consequences.

@@

The datasets for this report were obtained from the data section for "The Puzzle of Falling US Birth Rates since the Great Recession" [@citepaper], a paper from the Journal of Economic Perspectives [@citewebsite]. The data used by the paper was compiled by various sources such as the NCHS Natality Database [@citenchs], CDC SEER [@citecdcseer], the Current Population Survey Annual Social and Economic Supplement (CPS ASEC) [@citeipums], the New York Federal Reserve [@citefed], and the Integrated Public Use Microdata Series (IPUMS) [@citeipums]. The paper and datasets were last accessed on February 16th, 2022. The first two sources collect and estimate data via national census from US citizens and residents by county or state over the years 1968 to 2019. The birth rate data was obtained from a sample of 100 percent of birth certificates (i.e the population) in some states, and a sample of 50 percent of the birth certificate in other states, which were randomly sampled [@citenchs]. The national population data comes from the census, a long running non-voluntary survey funded by the government released on a predetermined cycle, and counts the entire population of the United States by county, thus response was mandatory by government mandate [@citecdcseer]. The datasets from the other sources compiled by the paper authors typically include data by year and by state for several demographics. These datasets were gotten from voluntary surveys so not everyone responded, thus the data was often recorded as 0, `NA`, or was estimated. The data is mainly numeric, counting the number of births, total population, or women in a particular economic/social/educational demographic, age group, or ethnicity. Meanwhile, categorical data mainly appeared in the form of state.

@@

For data cleaning, our target sample and focus was young mothers from two generations: millennials, who were between 15 to 24 years in 2000, and generation Z, who were between 15 to 24 years old in 2020. We observed birth rates from 2000 to 2020, in order to see the effect of relatively short-term effects on birth rates between the generations. Millennials are often considered to be born between 1980 to 1995, and generation Z from 1996 to 2012 [@citedimock]. We also considered birth rates for White, Black, and Hispanic women that were in our age ranges on interest, as we wanted to compare the birth rates between them. We thus filtered our datasets based on those criteria or used the closest criteria when they were not available, i.e., 2019 data is used if 2020 data is not available. We dropped any `NA` values from the dataset if they appeared.

@@

We included datasets and variables if they were considered important for answering the research question,  or if they revealed important information. We included variables on race, as this was needed to compare birth rates for different races. We also had variables on total number of births and birth rates, both of which were used to analyze the decline in births. We had a total population variable, used to get the birth rates and our demographics' share of the population. We had variables on birth rates for different age groups, states, and years, all of which were used to analyze the trends in birth rates for different groups.  We also included variables on government policy, rent costs, student debt, religious beliefs, and childcare costs. These variables were used to signify a possible relationship between them and birth rates. All the variables listed were numeric, with the exception of race and age group.

We created several of our own datasets by modifying and joining variables. One dataset was obtained by adding all the educational statuses for each age-race combination from another dataset to get the total birth rate for each combination, and then compiling the values by state and year. We also got a variable for total population by adding the total population for each state in a year.  We obtained the final dataset used to create the map of birth rate decline by combining various other datasets by year, and subtracting the birth rates of the later years by that of the earlier years. The final dataset that showed how our demographics of interest contributed to total birth rate decline, was created by joining several datasets which compared important information, then dividing the columns of our demographic by the column of total birth rates to get the relative contributions. The final dataset used to display the change in explanatory variables over time was obtained by combining the datasets, each of which explained the changes in a single variable. Most final datasets include either 51 observations (for each state), 102 observations (for each state by year), or 20 observations (for each year).

We modified many variables for the report. We created variables for total birth rates of each race-age combination by adding up all educational statuses for each race and age group combination and dividing by the total population, as each person whose birth rate was recorded also had their educational status recorded. We also modified age group variables by changing the columns of 15-19 and 20-24 age to rows within one column. We did the same for race variables, modifying the columns to be rows within one group, and then multiplying the values column by 10,000 and dividing by the total female population to get the birth rates for each race. These variables were used to get the trend in birth rates for our age demographics, and trend in birth rates in race for our age demographics. We obtained a variable which displayed the decline in birth rates from 2000 to 2019 by state, by dividing the number of births by population to get birth rates, grouping them by year and state, and then subtracting the values of the later year from the values of the earlier year.

We created a variable for the total number of births by adding up all the births by state and then grouping them by year. We also created a variable that stated the total population of each race-age combination by similar methods. We obtained a variable that got the percentage of population of a demographic by dividing the sum of that demographic by the total population and multiplying by 100. We produced a variable of contribution to total birth rate decline by subtracting 2019 birth rates of the age group from the 2000 birth rates of the same age group and dividing that by the total birthrate value for 2000, then multiplying by 200. We got the variables for all our explanatory factors by grouping them by year and getting the mean values from all the states. For the average debt, we multiplied by 10,000 as that is how it was shown in the original source. We got the variable on the importance of religion by dividing the number of people who felt religion was important over the total number surveys. Note that for these variables we needed to use 2004 data as 2000 data was not available.

@@

We see that for the 15 to 19 age group, the data is quasi-normal, with most births for the relevant years being between 60 to 190 thousand. The data seems evenly skewed and unimodal. We see that teenage births tend to be less popular, as the data seems slightly right skewed. This also means that the average number of births will be below the median number of births, which makes sense as teen births in some areas of the US are far higher than others [@Kearney2012]. We also see that Black teenage mothers are less common, as they have the lowest births for most years. This is most likely because they are a much smaller segment of the population than the others. However, we see that black mothers still have a moderate number of births, indicating that they are disproportionately appearing as teenage mothers. Hispanic teenage mothers are more middle ground, having between 60 to 100 thousand births per year. White teenage mothers are the most common, due to them being the largest in population. They tend to have most of their distribution towards the higher numbers.

For the 20-24 age group, there seems to be a slight bimodal distribution, with many births being on two extremes. Due to this it is difficult to determine whether the mean or the median of births will be larger.  We see that there are far more women giving birth at this age group than in the previous age group. We see that Black mothers are overwhelmingly on the lower end of births, which brings into question why teenage Black mothers had more prevalent births compared to their elder counterparts. Hispanic mothers also tend to be at the lower end of the scale but are more spread out. White births are the largest, taking the higher end of the spectrum, indicating their higher populations, or that white women are more likely to have children at these ages. We see that births for Black and Hispanic mothers at age 20-24 are only double the births of their teenage counterparts, while White mothers tend to be as much as 5 times the number of their teenage counterparts, Therefore, teenage motherhood in Black and hispanic women may be more common.

In contrast, in (Figure \@ref(fig:figthree)), the distribution for the total population of women for the past 20 years seems uniformal, with almost all years having their own value. This is because the population of a nation either increases or decreases, rarely staying the same. Due to this, we can say that the mean and median population are approximately the same. However there are some years which had around the same population, this could be due to declining population as a later year has the same population as an earlier year. From the scatter plot we see that the total female population has been increasing in recent years, with a seemingly exponential increase. This increase in the population could be due to increased amounts of immigration and the increased lifespan of the elderly to offset the decreasing birth rates. However, the two dips in the scatter plot correspond to the previous histogram of which had years with similar values.

@@

(Figure \@ref(fig:figfour)) shows the trend of births for the two age groups in a 20 year period. We see that there has been a significant decrease in the birth rates, especially in the 15-19 age group, dropping to half of its 2000 levels. This indicates a decrease in teenage pregnancy nationwide. The birth rates for the 20-24 age dropped by around 40% over the same period. This may be because generation Z women are deciding to have children later in life. There was a baby boom in 2006 [@citecatholic], which is the reason for the slight increase in birth rates around that time. However overall birth rates for young women have dropped.  According to [@Caldwell2006], there is little reason to believe these birth rates will increase any time soon. There are some possible explanations behind this phenomenon, including better sex education for teenagers leading to a decrease in teen pregnancy, more women focusing on their careers and thus postponing births [@citeexecutive], or the increased acceptance of abortions. These declining birth rates could also be the cause of declining population. There are also economic factors correlated to this decline, including the rising cost of living.

```{r figfour, out.width="90%", fig.cap = "Trend in birth rates for the 15-19 and 20-24 age groups from 2000 to 2020", echo = FALSE}

@@

(Figure \@ref(fig:figfive)) shows the trend of births for different races of women from the ages of 15 to 24 over the past 20 years. We see that the birthrate for Hispanic mothers significantly decreased, going from about 63 births per 1000 population to just about 40 births for 1000 population. This can either be because the population of Hispanic women has increased due to immigration, or that the number of births has decreased. The birth rates for Black women have also halved, going from 50 births per 1000 to just 25 births per 1000. The White birth rates however have decreased the most, going from about 125 births to about 65 births through the 20 years. We see that millennial mothers (mothers in 2000) had a large difference in the birth rates for different races. However, for generation Z they have converged more closely due to the decrease in birth rates for all races. This could be due to economic factors directly related to race, such as inequality in the health care system [@Davis2018]. As the income for Hispanic families increases and they become more westernized, it is possible that they may be adhering to the western standard of fewer children [@immigbirth].

```{r figfive, out.width="90%", fig.cap = "Trend in birth rates by race from 2000 to 2020", echo = FALSE}

@@

(Table \@ref(tab:tabone)) shows the contribution of the demographics of interest to the total decline in birth rates from 2000 to 2019. It also shows each demographic's share of the total female population for those years. We see that white women in both age groups have contributed the most to the decline, with a total of 18.2% of the total decrease. White 15 to 19 year old women have caused a 7% drop in the total birth rate, indicating a rapid decline in the teenage pregnancy for this race [@Wingo2011]. White 20 to 24 year old women have contributed the most to the decline, possibly due to economic and cultural factors over the past 20 years [@Maher2007]. Black women for both age groups have not shown a significant contribution to the decline in birth rates over this period however, either because teenage pregnancy rates for this race have not decreased significantly or because they are a much lower percentage of the population. Black women are disproportionately poorer [@citepovertyrace], and low economic status leads to higher teenage pregnancy rates [@Gold2001]. The contribution of Hispanic women of ages 20 to 24 is less than their teenage counterparts, showing that there is not a large decrease in their birth rates as opposed to white women. We see that White womens' share of the population from 2000 to 2019 has decreased due to the large decrease in birth rates. However Black and Hispanic womens' share of the population over this period has either increased or not changed, likely due to increasing immigration from these two races over the past 20 years.

@@

# getting each race - age combination's total contribution to declining birth rates
contribution = birthrates %>%
  pivot_wider(names_from = year, values_from = birthratevalue)
contribution$contribution = round(((contribution$"2000" - contribution$"2019")/65.9)*200,1)

# getting the percentage of the population
percentpop = share_of_pop %>%
  pivot_wider(names_from = year, values_from = value)
percentpop$"2000" = round(percentpop$"2000"*100,1)
percentpop$"2019"= round(percentpop$"2019"*100,1)

@@

(Figure \@ref(fig:figsix)) shows the number of births for those aged 15-24 from 2000 to 2019, according to the year's unemployment rate. We notice that the graph generally has a decreasing pattern in terms of the average number of births per year in the United States which is in line with what we observed earlier. Between 2004 and 2005, the birth rate for the 15-19 group continued to decrease whereas the birth rate for the 20-24 age group began to increase. We see that the unemployment rate is low in this period. According to [@citecatholic], in 2006, the United States had their highest number of births than in the last 45 years prior, causing a baby boom. However, after this, the birth rate began to quickly decline and unemployment rates were high, reaching 8% by 2008 and continuing until 2012. This is likely due to the great recession that occurred during this period, such that it is possible women could not afford bringing new children. The birth rate, while still decreasing for both age groups, becomes less severe from 2012 onward, and we see that unemployment reduces. The initial conclusion made by this is that low unemployment rates are related to a higher number of births and higher unemployment rates are related to a lower number of births. However, beyond 2014, despite the low unemployment rate, we see the birth rate continues to decrease. We note that the unemployment rate is more likely to affect women in the 20-24 age group compared to their younger counterparts, as they are more involved in the workforce.

```{r figsix, out.width="90%", fig.cap = "Average birth Rates against level of unemployment for the 15 to 19 and 20 to 24 age groups from 2000 to 2020", echo = FALSE}

@@

# creating scatter plot for the 15 to 19 age group
g1519 <- graph %>% ggplot(mapping = aes(x = year, y = brate_1519, color = Unemployment)) + geom_point() + theme_minimal() + scale_x_continuous(breaks = seq(1980,2020,by = 2))+ labs(x = "Year", y = "Birth Rate for Ages 15-19") + 
  labs(color= "Unemployment Rate (%)")

# creating scatter plot for the 20 to 24 age group
g2024 <- graph %>% ggplot(mapping = aes(x = year, y = brate_2024, color = Unemployment)) + geom_point() + theme_minimal() + scale_x_continuous(breaks = seq(1980,2020,by = 2))+ labs(x= "Year", y = "Birth Rate for Ages 20-24") +
  labs(color= "Unemployment Rate (%)")

@@

(Figure \@ref(fig:figseven)) shows the percentage decline in birth rates for young women, specifically for the 20-34 age group, by state from 2000 to 2019. We see that throughout the U.S, there has been a decline in birth rates in most states, with some states having a more severe decline than others. We see that states on the west coast tend to have a larger decline than states on the east coast, with California having a very high decline in birth rates, and North Dakota having relatively little change. This could be attributed to the west coast being more liberal [@citeleft] and thus less adhering to gender roles, such that women have children later or not at all. In comparison, the east coast, midwestern and southern states have less severe changes in their birth rates, with some having no change or a slight increase in birth rates. Coincidentally, these states tend to be more conservative or religious, so they accept gender norms and refusal of new changes in culture. The exception seems to be Utah, which is conservative and religious due to Mormonism but has the largest decline in birth rates. However, the general trend still holds, and as some states are more costly to live in than others, this could also explain the differences.

@@

# mean rent by year, actually recorded monthly
renttable = rent %>%
  group_by(year) %>%
  summarize(meanrent = mean(avg_grossrent)) %>%
  mutate(meanrent = round(meanrent,1))
